Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-27 paper/current/main.tex
small edits. move network centrlaity to SI

Human mobility and migration drives major social phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Common models of mobility, such as the Gravity Model~\autocite{zipf1946gravity} aim to capture mobility between places using their population and the \textit{distance} between them.
However, geographic distance is less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} wherein the physical barrier shrinks while linguistic, economic, cultural, and historical similarities matter more.
Alternative notions of functional distance~\autocite{brown1970migration, brown1970functional, kim2018functional} attempt to re-define distance by using some of aforementioned factors~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional} but are often arbitrary and consider only single isolated facets.
Here, using a rich dataset of scientific mobility, we demonstrate that a data-driven neural embedding approach~\autocite{mikolov2013word2vec} can successfully encodes a functional distance between organizations and captures the latent and multi-faceted structure of scientist's mobility.
We demonstrate that the cosine distance between organization embedding vectors, constructed using the affiliation trajectories of scientists in the Web of Science database, better explain and predict real mobility than geographic distance.
Exploiting their unique properties, we demonstrate that the embedding space encodes not only geographic relations between organizations, but also relations based on culture, language, and academic prestige at multiple levels of granularity.
Also captured in the embedding space is a universal pattern capturing the relationship between organization's size and their connectedness.
The ability to learn scalable, dense, and meaningful high-resolution vector representations of mobility directly from the data can open up a new avenue of studying human mobility.

@@

The embedding space encodes the prestige hierarchy of U.S. universities well enough to re-construct university rankings. 
We derive a ranking for universities using the SemAxis prestige axis and compare them to traditional university rankings.
The embedding-based ranking is defined from the position of organizations along the prestige axis, with the top-ranked being the nearest to the high-ranked pole.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the embedding-based ranking achieves correlation with the Times Ranking comparable to that between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a well-known bibliometrically-informed university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding based ranking and the Times ranking is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities are significantly correlated with the Times Ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
The strongest outliers that were ranked more highly in the Times Ranking than the embedding based ranking tend to be large state universities such as Arizona State University and  University of Florida;
Those ranked higher in the embedding based ranking tend to be relatively-small universities such as the University of San Francisco and University of Maryland Baltimore County near major urban areas.
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility with a high degree of resolution and reliability, even without providing any explicit information on prestige~\autocite{deville2014career, clauset2015hierarchy}.

The performance of the embedding ranking in reconstructing the Times ranking is comparable to that of network-derived measures such as weighted degree (Spearman's $\rho = 0.73$, Fig.~\ref{fig:supp:centrality_times_compare}a) and eigencentrality centrality (Spearman's $\rho = 0.76$, Fig.~\ref{fig:supp:centrality_times_compare}b).
However, while both embedding- and network-based measures relate to university prestige, they are qualitatively and quantitatively different.
The embedding-ranking of U.S. universities is less correlated with weighted degree (Spearman's $\rho = 45$, Fig.~\ref{fig:supp:centrality_semaxis_compare}a) and eigenvector  centrality (Spearman's $\rho = 0.55$) than with the Times Ranking itself (Spearman's $\rho = 0.73$, Fig.~\ref{fig:supp:centrality_semaxis_compare}b). 
The embedding ranking over-ranks large research-intensive universities such as North Carolina State University, University of Florida, and Texas A\&M University, whereas the network-derived ranking over-ranks smaller, more specialized universities such as Brandeis University, Yeshiva University, and University of San Francisco. 
This suggests that the embedding encodes encodes information on prestige hierarchy at least as well as a network representation, but that the information encoded is something different from centrality.

The axes can further be visualized to examine the the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts that aids visualization.
Prestigious universities such as Colombia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reasoning about the prestige of organizations typically excluded from rankings such as government organizations (Fig.~\ref{fig:semaxis}d), research institutes (Fig.~\ref{fig:semaxis}e), and teaching colleges (Fig.~\ref{fig:semaxis}f).

@@

>>>>

Human mobility and migration drives major social phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Common models of mobility, such as the Gravity Model~\autocite{zipf1946gravity} aim to capture mobility between places using their population and the \textit{distance} between them.
However, geographic distance is less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} wherein the physical barrier shrinks while linguistic, economic, cultural, and historical similarities matter more.
Alternative notions of functional distance~\autocite{brown1970migration, brown1970functional, kim2018functional} attempt to re-define distance by using some of aforementioned factors~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional} but are often arbitrary and consider only single isolated facets.
Here, using a rich dataset of scientific mobility, we demonstrate that a data-driven neural embedding~\autocite{mikolov2013word2vec} can successfully encodes a functional distance between organizations and captures the latent and multi-faceted structure of scientist's mobility.
We demonstrate that the cosine distance between organization embedding vectors, constructed using the affiliation trajectories of scientists in the Web of Science database, better explain and predict real mobility than geographic distance.
Exploiting their unique properties, we demonstrate that the embedding space encodes not only geographic relations between organizations, but also relations based on culture, language, and academic prestige at multiple levels of granularity.
Also captured in the embedding space is a universal pattern capturing the relationship between organization's size and their connectedness.
The ability to learn scalable, dense, and meaningful high-resolution vector representations of mobility directly from the data can open up a new avenue of studying human mobility.

@@

The embedding space encodes the prestige hierarchy of U.S. universities well enough to re-construct university rankings. 
We derive a ranking for universities using the SemAxis prestige axis and compare them to traditional university rankings.
The embedding-based ranking is defined from the position of organizations along the prestige axis, with the top-ranked being the nearest to the high-ranked pole.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the embedding-based ranking achieves correlation with the Times Ranking comparable to that between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a well-known bibliometrically-informed university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding based ranking and the Times ranking is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities are significantly correlated with the Times Ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
Performance is also comparable to network-driven measures such as degree strength (Spearman's $\rho = 0.73$) and eigenvector centrality (Spearman's $\rho = 0.76$, see Supporting Information). 
The strongest outliers that were ranked more highly in the Times Ranking than the embedding based ranking tend to be large state universities such as Arizona State University and  University of Florida;
Those ranked higher in the embedding based ranking tend to be relatively-small universities such as the University of San Francisco and University of Maryland Baltimore County near major urban areas.
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility with a high degree of resolution and reliability, even without providing any explicit information on prestige~\autocite{deville2014career, clauset2015hierarchy}.

The axes can further be visualized to examine the the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts that aids visualization.
Prestigious universities such as Colombia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reasoning about the prestige of organizations which are typically not ranked, such as government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e), or for which rankings are often low-resolution and idiosyncratic, such as for regional and liberal arts universities (Fig.~\ref{fig:semaxis}f).

@@

%
% S7 TEXT
\paragraph*{S7 Text}
\label{si:text:network_ranking}
{\bf Reconstructing Times ranking with network measure.}

The performance of the embedding ranking in reconstructing the Times ranking is comparable to that of network-derived measures such as degree strength (Spearman's $\rho = 0.73$, Fig.~\ref{fig:supp:centrality_times_compare}a) and eigencentrality centrality (Spearman's $\rho = 0.76$, Fig.~\ref{fig:supp:centrality_times_compare}b).
However, while both embedding- and network-based measures relate to university prestige, they are qualitatively and quantitatively different.
The embedding-ranking of U.S. universities is less correlated with weighted degree (Spearman's $\rho = 0.45$, Fig.~\ref{fig:supp:centrality_semaxis_compare}a) and eigenvector  centrality (Spearman's $\rho = 0.55$) than with the Times Ranking itself (Spearman's $\rho = 0.73$, Fig.~\ref{fig:supp:centrality_semaxis_compare}b). 
The embedding ranking over-ranks large research-intensive universities such as North Carolina State University, University of Florida, and Texas A\&M University, whereas the network-derived ranking over-ranks smaller, more specialized universities such as Brandeis University, Yeshiva University, and University of San Francisco. 
This suggests that the embedding encodes information on prestige hierarchy at least as well as a network representation, but that the information encoded is something different from centrality
